---
title: "Atlanta United Project 2"
author: "Jake Federman"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(fmsb)
library(scales)
library(kableExtra)
options(scipen = 999)
```

Objective: Find 3 attacking players who stand out as potential targets for MLS clubs

#Reading in the Data
```{r}
events <- readRDS("atlutd_datascientist_project2_eventdata.rds")
schedule2 <- read_csv(file = "atlutd_datascientist_project2_schedule.csv")
caps <- read_csv(file = "atlutd_datascientist_project2_player_mins_appearances.csv")
```

## Filter the data down to only attacking player events

```{r}
events <-
  events %>% 
  filter(position.name %in% c("Center Forward",
                              "Center Attacking Midfield",
                              "Right Attacking Midfield",
                              "Left Attacking Midfield",
                              "Right Wing",
                              "Left Wing",
                              "Right Center Forward",
                              "Left Center Forward"))
  
```


# Creating new stats

## Passes Completed Over Expected

We will create an Outs Above Average style metric for pass completion.

Let $X_{ij}$ be the probability of completing pass $i$ by player $j$.
Let $Y_{ij}$ be the outcome of pass i by player j: 0 if incomplete, 1 if complete
Let $n$ be the number of passes player $j$ attempted this season.

Player $j$'s $PCOE$ (passes completed over expected) is equal to the sum of his $Y_i$ - $X_i$ values

We will not include set pieces 

```{r}
passes <-
  events %>% 
  filter(type.name == "Pass" & 
           (is.na(pass.type.name) |
            (!pass.type.name %in% 
               c("Kick Off", "Corner", "Free Kick", "Throw-in")))) %>%
  mutate(
    pass_complete = case_when(
    is.na(pass.outcome.name) ~ 1,
    TRUE ~ 0),
    pcoe = pass_complete - pass.pass_success_probability
    ) %>% 
  select(player.id, pass_complete, pass.pass_success_probability, pcoe)

pcoe_leaderboard <-
  passes %>% 
  group_by(player.id) %>% 
  summarize(num_passes = n(),
            pcoe_total = sum(pcoe, na.rm = TRUE),
            pcoe_per_pass = pcoe_total/num_passes) %>%
  arrange(desc(pcoe_total))

#Standardize
standardize <- function(x) {
  mu <- mean(x, na.rm = TRUE)
  sigma <- sd(x, na.rm = TRUE)
  return( (x - mu)/sigma )
}

z_pcoe_leaderboard <-
  pcoe_leaderboard %>% 
  mutate(z_pcoe_total = standardize(pcoe_total),
         z_pcoe_per_pass = standardize(pcoe_per_pass)) %>% 
  select(1, 2, 5, 6)

z_pcoe_per_pass_leaders <-
  z_pcoe_leaderboard %>% 
  arrange(desc(z_pcoe_per_pass))

pcoe_leaderboard %>% 
  select(player.id, num_passes, pcoe_total) %>%
  arrange(desc(pcoe_total)) %>% 
  head(10) %>% 
  kable(format = "latex", booktabs = TRUE, digits = 3,
        col.names = c("Player", "Passes", "Total PCOE"),
        caption = "Passes Completed Over Expected Leaders",
        align = c("l", "c", "c")) %>% 
  kable_styling(position = "center", latex_options = "HOLD_position")

pcoe_leaderboard %>% 
  select(player.id, num_passes, pcoe_per_pass) %>%
  arrange(desc(pcoe_per_pass)) %>% 
  head(10) %>% 
  kable(format = "latex", booktabs = TRUE, digits = 3,
        col.names = c("Player", "Passes", "PCOE per Pass"),
        caption = "Passes Completed Over Expected (per Pass) Leaders",
        align = c("l", "c", "c")) %>% 
  kable_styling(position = "center", latex_options = "HOLD_position")
```

## Aggregations Involving Expected Goals

```{r}
shots <-
  events %>% 
  filter(type.name == "Shot" & shot.type.name == "Open Play") %>% 
  select(player.id, shot.outcome.name, shot.statsbomb_xg, shot.shot_execution_xg,
         shot.shot_execution_xg_uplift)
```

```{r}
#Pre-shot xG leaders
pre_shot_xg_leaders <-
  shots %>% 
  group_by(player.id) %>% 
  summarize(num_shots = n(),
            total_pre_shot_xg = sum(shot.statsbomb_xg, na.rm = TRUE),
            pre_shot_xg_per_shot = total_pre_shot_xg/num_shots) %>% 
  arrange(desc(total_pre_shot_xg))

#Post shot xG leaders
post_shot_xg_leaders <-
  shots %>% 
  group_by(player.id) %>% 
  summarize(num_shots = n(),
            total_post_shot_xg = sum(shot.shot_execution_xg, na.rm = TRUE),
            post_shot_xg_per_shot = 
              total_post_shot_xg/num_shots) %>% 
  arrange(desc(total_post_shot_xg))

#xG above expected (post shot xG - pre shot xG)
xgoe_leaders <-
  shots %>% 
  group_by(player.id) %>% 
  summarize(num_shots = n(),
            total_xgoe = sum(shot.shot_execution_xg_uplift, na.rm = TRUE),
            xgoe_per_shot = total_xgoe/num_shots) %>% 
  arrange(desc(total_xgoe))
```

## 50/50s

```{r}
fifty_fifties <-
  events %>% 
  filter(type.name == '50/50') %>% 
  select(player.id, `50_50.outcome.id`) %>% 
  mutate(fifty_win = case_when(
    `50_50.outcome.id` <= 2 ~ 0,
    TRUE ~ 1
  )) %>% 
  group_by(player.id) %>% 
  summarize(`50_50s` = n(),
            wins_50_50 = sum(fifty_win, na.rm = TRUE),
            expected_wins_50_50 = `50_50s`/2,
            wins_above_expected_50_50 = wins_50_50 - expected_wins_50_50) %>% 
  ungroup()
```

## Dribbles

```{r}
dribbles <-
  events %>% 
  filter(type.name == "Dribble") %>% 
  mutate(dribble_success = case_when(
    dribble.outcome.id == 8 ~ 1,
    TRUE ~ 0
  )) %>% 
  select(player.id, dribble_success) %>% 
  mutate(overall_success_rate = sum(dribble_success)/n()) %>% 
  group_by(player.id, overall_success_rate) %>% 
  summarize(num_dribble_attempts = n(),
            num_dribble_successes = sum(dribble_success, na.rm = TRUE),
            dribble_success_rate = num_dribble_successes/num_dribble_attempts) %>% 
  ungroup() %>% 
  mutate(expected_dribble_successes = overall_success_rate * num_dribble_attempts,
         dribbles_over_expected = 
           num_dribble_successes - expected_dribble_successes) %>% 
  select(-overall_success_rate)
```

## Join the new stats together

```{r}
new_stats_leaderboard <-
  pcoe_leaderboard %>% 
  inner_join(pre_shot_xg_leaders, by = 'player.id') %>% 
  inner_join(post_shot_xg_leaders, by = 'player.id') %>% 
  inner_join(xgoe_leaders, by = 'player.id') %>% 
  inner_join(dribbles, by = 'player.id') %>% 
  left_join(fifty_fifties, by = 'player.id') %>% 
  mutate(across(everything(), ~replace_na(., 0)))

new_stats_totals <-
  new_stats_leaderboard %>% 
  select(player.id, num_passes, num_shots, num_dribble_attempts,
         `50_50s`, pcoe_total, total_pre_shot_xg,
         total_xgoe, dribbles_over_expected, wins_above_expected_50_50)

```

## Combine new leaderboard with minutes played data
```{r}
caps_simple <-
  caps %>% 
  mutate(minutes_per_game = player_season_minutes/player_season_appearances) %>% 
  select(player_id, player_season_minutes, minutes_per_game) %>% 
  rename(player.id = player_id)

combined_stats <-
  new_stats_totals %>% 
  inner_join(caps_simple, by = 'player.id')
```

## Put key stats on a per 90 minute basis
```{r}
combined_stats_per_90 <-
  combined_stats %>% 
  mutate(pcoe_per_90 = 90*(pcoe_total/player_season_minutes),
         xg_per_90 = 90*(total_pre_shot_xg/player_season_minutes),
         xgoe_per_90 = 90*(total_xgoe/player_season_minutes),
         doe_per_90 = 90*(dribbles_over_expected/player_season_minutes),
         wae_50_50 = 90*(wins_above_expected_50_50/player_season_minutes)) %>% 
  select(-c(6:10))
```

```{r}
#Take z-scores
z_combined_stats_per_90 <-
  combined_stats_per_90 %>% 
  mutate(z_pcoe_per_90 = standardize(pcoe_per_90),
         z_xg_per_90 = standardize(xg_per_90),
         z_xgoe_per_90 = standardize(xgoe_per_90),
         z_doe_per_90 = standardize(doe_per_90),
         z_wae_50_50 = standardize(wae_50_50)) %>% 
  select(1:7, 13:17)

z_sum_combined_stats_per_90 <-
  z_combined_stats_per_90 %>% 
  group_by(player.id) %>% 
  mutate(total_z_score = sum(across(c(starts_with("z_"))))) %>% 
  arrange(desc(total_z_score))
```

# The best players by total_z_score
```{r}
z_sum_combined_stats_per_90 %>% 
  select(player.id, total_z_score) %>% 
  head(10) %>% 
  kable(format = "latex", booktabs = TRUE, digits = 3,
        col.names = c("Player", "Total Z-Score"),
        caption = "The best players by total z score",
        align = c("l", "c", "c")) %>% 
  kable_styling(position = "center", latex_options = "HOLD_position")
```

# Hand Picking the Ideal Targets

I will hand pick the ideal targets taking into account not only total z-score, but also sample size. The top player according to total z score has only played 75 minutes all season across 2 matches, and he has 2 shots. We should not take a chance on him with this little information.

## Player 1: Player 4269
```{r, fig.cap="Radar plot for Target 1, player 4269"}
row_4269 <- 
  z_sum_combined_stats_per_90 %>% 
  filter(player.id == 4269)

radar_values <- 
  row_4269 %>%
  ungroup() %>% 
  select(z_pcoe_per_90, z_xg_per_90, z_xgoe_per_90, z_doe_per_90, z_wae_50_50) %>% 
  rename(`Passing (-0.09)` = z_pcoe_per_90,
         `Scoring chances (2.31)               ` = z_xg_per_90,
         `Finishing (1.91)` = z_xgoe_per_90,
         `Dribbling (1.08)` = z_doe_per_90,
         `50/50s (0.2)` = z_wae_50_50)

rescaled <- 
  as.data.frame(lapply(radar_values, function(x) (x + 3) / 6))

radar_df <-
  rbind(rep(1, 5),
                  rep(0, 5),
                  rescaled)

colnames(radar_df) <- 
  colnames(radar_values)
rownames(radar_df) <-
  c("max", "min", "Player")

radarchart(radar_df,
           pcol = "blue",
           pfcol = alpha("blue", 0.4),
           plwd = 2,
           cglcol = "grey",
           cglty = 1,
           axislabcol = "grey20",
           vlcex = 0.8,
           title = "Player 4269 Radar Plot")

title(sub = "Total z-score = 5.41 (2nd)", cex.sub = 1)


```

## Scouting for the second and third players:

In searching for suitable transfer targets, we must look for both good quality and great value. As a result, for the next two targets, I will not merely choose the best players, but I will look for ones that are being used as substitutes. If we were to inquire about a player who is playing 95 minutes every match, the team will likely demand a high price, as that player is a crucial part of their organization. Perhaps they will be more willing to sell a player who is on the pitch less, maybe coming off the bench as a sub, allowing us to get a better price.

```{r}
z_sum_combined_stats_per_90 %>% 
  filter(minutes_per_game <= 50 & player_season_minutes >= 250
         & num_passes >= 50 & num_shots >= 5 & total_z_score >= 1.5)
```

## Player 2: Player 52065
```{r,fig.cap="Radar plot for Target 2, player 52065"}
row_52065 <- 
  z_sum_combined_stats_per_90 %>% 
  filter(player.id == 52065)

radar_values <- 
  row_52065 %>%
  ungroup() %>% 
  select(z_pcoe_per_90, z_xg_per_90, z_xgoe_per_90, z_doe_per_90, z_wae_50_50) %>% 
  rename(`Passing (1.29)` = z_pcoe_per_90,
         `Scoring chances (1.04)               ` = z_xg_per_90,
         `Finishing (0.329)` = z_xgoe_per_90,
         `Dribbling (0.582)` = z_doe_per_90,
         `50/50s (-0.784)` = z_wae_50_50)

rescaled <- 
  as.data.frame(lapply(radar_values, function(x) (x + 3) / 6))

radar_df <-
  rbind(rep(1, 5),
                  rep(0, 5),
                  rescaled)

colnames(radar_df) <- 
  colnames(radar_values)
rownames(radar_df) <-
  c("max", "min", "Player")

radarchart(radar_df,
           pcol = "blue",
           pfcol = alpha("blue", 0.4),
           plwd = 2,
           cglcol = "grey",
           cglty = 1,
           axislabcol = "grey20",
           vlcex = 0.8,
           title = "Player 52065 Radar Plot")

title(sub = "Total z-score = 2.45 (1st among players with more than 200 min
      on the season and less than 45 minutes per game)", cex.sub = 1)


```

## Player 3: Player 42137
```{r,fig.cap="Radar plot for Target 3, player 42137"}
row_42137 <- 
  z_sum_combined_stats_per_90 %>% 
  filter(player.id == 42137)

radar_values <- 
  row_42137 %>%
  ungroup() %>% 
  select(z_pcoe_per_90, z_xg_per_90, z_xgoe_per_90, z_doe_per_90, z_wae_50_50) %>% 
  rename(`Passing (0.390)` = z_pcoe_per_90,
         `Scoring chances (-0.212)               ` = z_xg_per_90,
         `Finishing (1.71)` = z_xgoe_per_90,
         `Dribbling (0.019)` = z_doe_per_90,
         `50/50s (0.196)` = z_wae_50_50)

rescaled <- 
  as.data.frame(lapply(radar_values, function(x) (x + 3) / 6))

radar_df <-
  rbind(rep(1, 5),
                  rep(0, 5),
                  rescaled)

colnames(radar_df) <- 
  colnames(radar_values)
rownames(radar_df) <-
  c("max", "min", "Player")

radarchart(radar_df,
           pcol = "blue",
           pfcol = alpha("blue", 0.4),
           plwd = 2,
           cglcol = "grey",
           cglty = 1,
           axislabcol = "grey20",
           vlcex = 0.8,
           title = "Player 42137 Radar Plot")

title(sub = "Total z-score = 2.10 (2nd among players with more than 200 min
      on the season and less than 45 minutes per game)", cex.sub = 1)

```

### Extra PCOE leaderboard
```{r}
combined_stats_per_90 %>% 
  select(player.id, pcoe_per_90) %>% 
  arrange(desc(pcoe_per_90)) %>% 
  head(10) %>% 
  kable(format = "latex", booktabs = TRUE, digits = 3,
        col.names = c("Player", "PCOE per 90 Minutes"),
        caption = "PCOE per 90 Minutes Leaderboard",
        align = c("l", "c")) %>% 
  kable_styling(position = "center", latex_options = "HOLD_position")
```

